%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Medium Length Professional CV
% LaTeX Template
% Version 2.0 (8/5/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Trey Hunner (http://www.treyhunner.com/)
%
% Important note:
% This template requires the resume.cls file to be in the same directory as the
% .tex file. The resume.cls file provides the resume style used for structuring the
% document.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\newcommand{\latex}{\LaTeX\xspace}
\newcommand{\tex}{\TeX\xspace}

\documentclass{resume}

\usepackage[left=0.5in,top=0.4in,right=0.5in,bottom=0.4in]{geometry} % Document margins
\usepackage{fontawesome5}

\name{William Gertler} % Your name
\address{(506)~$\cdot$~259~$\cdot$~8831 \\ will@wgertler.com \\ wgertler.com \\ \faGithubSquare\ swickrotation} % Your phone number and email and website

\begin{document}

%----------------------------------------------------------------------------------------
%	WORK EXPERIENCE SECTION
%----------------------------------------------------------------------------------------

\begin{rSection}{Experience}
    \begin{rSubsection}{Manager and Lead Bartender}{June 2018 - May 2024}{Loloan Lobby Bar}{Waterloo, ON}
        \item Created the very best cocktail program in the region.
        \item Raised bar section revenue by 18\% and total revenue by 12\%.
        \item Led a bar team (3-5 persons) that delivered award-winning service six nights a week.
        \item Developed a wine-pairing system gaining a place on vineroutes' list of best restaurants in Ontario.
    \end{rSubsection}

    \begin{rSubsection}{Quantum Technologies Consultant}{March 2019 - April 2020}{Chosen Plaintext, Inc.}{Waterloo, ON}
    \item Due diligence for investors, quality assurance of products and methodologies including stress-testing of products and facility inspections.
    \item Preparation of explanatory documents, slide decks, and analysis of whitepapers vital to informed decision-making on the venture-capital side.
    \end{rSubsection}

    \begin{rSubsection}{Optical Physicist, AR\&D}{April 2016 - September 2016}{North (acq. Google, 2020)}{Kitchener, ON}
        \item Achieved world-first breakthrough in hologram technology and optical fidelity.
        \item Key data acquisition and analytics leading to the development of Focals smartglasses.
    \end{rSubsection}

    \begin{rSubsection}{Research Assistant}{February 2014 - April 2015}{Femtosecond Laser Lab, Strickland-Sanderson group (Nobel Prize in Physics, 2018)}{Waterloo, ON}
        \item Coulomb Explosion Imaging Experiment developing environmentally sustainable and economically efficient graphene production.
        \item Ground-up fabrication of high-energy lasers, hardware and software development and maintenance. 
    \end{rSubsection}

\end{rSection}

%----------------------------------------------------------------------------------------
%	EDUCATION SECTION
%----------------------------------------------------------------------------------------

\begin{rSection}{Education}

    {\bf University of Waterloo} \hfill {\it Waterloo, ON} \vspace{2px}\\ 
    Ph.D. Applied Mathematics (Dissertation Pending)\hfill September 2018 - January 2023 \\ 
    B.Sc. Mathematical Physics, Honours \hfill September 2013 - April 2018\\
    %\small{Minor in Pure Mathematics} \\
    %\small{Specialization in Astrophysics}\\
    %\emph{Cumulative Average: 94.7\% }

\end{rSection}


%-------------------------------------------------------------
% SERVICE	
%-------------------------------------------------------------
\begin{rSection}{Service}
    \begin{rSubsection}
    {President}{September 2016-December 2016}{Undergraduate Physics Society of Waterloo}{Waterloo, ON}
\item Represented the physics department on the University of Waterloo's science faculty board, implemented a new constitution, coordinated events for 300+ students bringing in revenue in excess of 300\% of the student society's budget .
    \end{rSubsection}
\end{rSection}

%----------------------------------------------------------------------------------------
%	TECHNICAL STRENGTHS SECTION
%----------------------------------------------------------------------------------------

\begin{rSection}{Technical Skills}

\begin{tabular}{ @{} >{\bfseries}l @{\hspace{6ex}} l }
    Programming Languages & Python, MATLAB, HTML+CSS, Bash\\
%    Languages (Human) & English (Fluent)\\
Technology & Linux operating systems and development context, Git, Microsoft Office
\end{tabular}
\end{rSection}

%----------------------------------------------------------------------------------------
%	Awards & Scholarships
%----------------------------------------------------------------------------------------

%\begin{rSection}{Awards and Scholarships}
%    {University of Waterloo President's Scholarship} \hfill {\em September 2013} \\
%\end{rSection}

%-------------------------------------------------------------
%   TALKS	
%-------------------------------------------------------------
%\begin{rSection}{Invited Talks}
%    {Ethics in the Age of Lonely Qubits} \hfill {Alt-Tab, July 2019}\\
%    {Quantum Information Theory for the Gifted Amateur} \hfill {USPAM, November 2018}\\
%    {What is Quantum Theory and Why Should I Care?} \hfill {UWCSC, July 2018}\\
%\end{rSection}
%
%----------------------------------------------------------------------------------------

\end{document}
